A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-18 text/informal-overview.tex
FInish revising Seciton 1.
Reivised for easy comments.

\section{Informal Overview of \LMD}

@@

\red{\LTP\cite{Hanada2014}が頻発するが同一論文の複数回引用に何かルールはあるのか?}
\AI{ないです．たしかにあんまりつけるとうるさいですね．2.1 の冒頭とかは文ごと不要なのでは？}
\subsection{\LTP}

@@

\LTP\cite{Hanada2014} is a multi-stage calculus with CSP by Hanada and Igarashi.
In \LTP, brackets and escape are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$.
The type of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.
% Please notice that if $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for brackets and escape.
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M 
\end{align*}
It means escape cancels brackets.
% This reduction is called $\longrightarrow_\Lambda$ in \LMD.
% \red{この文はここに来るべきなのか? LTPの説明ではないが}

% transition variable and transition

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{transition variable} and
a sequence of transition variables is called \textit{transition}.
The empty transition is represented by $\varepsilon$.
They are used to show the thickness of brackets. \AI{Perhaps ``thickness'' doesn't make sense.}
For example, $\TB_\alpha (\lambda x:\I.x+10)$ is the fragment of code which becomes $(\lambda x:\I.x+10)$ after it was run once and
$\TB_\alpha \TB_\beta (\lambda x:\I.x+10)$, which is abbreviated as $\TB_{\alpha\beta} (\lambda x:\I.x+10)$,
becomes $(\lambda x:\I.x+10)$ after it was run it twice.
\red{run twiceは表現が微妙}

% transition abstraction and application

There are abstractions for transition variables and applications for transition abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M A$, respectively.
A transition abstraction binds a transition variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
It is only natural \AI{?} there is a reduction rule for transition application in \LTP.
The rule is the following.
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
For example, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ (\beta\gamma)$ reduces to $\TB_{\beta\gamma} (\lambda x:\I.x)$.

% transition-related symbols disappear when the empty transition is substituted.
% In short, about "run"

Another important rule about a transition variable is 
that symbols with transition variables disappear 
when the empty transition $\varepsilon$ is substituted to the transition variable.
The purpose of this rule is express \verb|run| without any special symbol.
In multi-stage calculus, \verb|run| is a very important operator which changes quoted code to the original code.
For example, $(\TB_\varepsilon (\lambda x:\I.x))$ is equivalent to $(\lambda x:\I.x)$.
Therefore, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ \varepsilon$ becomes $\lambda x:\I.x$
In \LTP, \verb|run| is realized with application to the empty transition $\varepsilon$.

% transition in typing rules

A type judgement of \LTP is of the form $\G \vdash M : \tau @ A$.
A transition in a judgement represents where the judgement is true.
For example, $\G \vdash (\lambda x:\I.x) : \I \to \I @ \alpha$ means 
term $\lambda x:\I.x$ has type $\I \to \I$ at transition $\alpha$.
Especially, terms without quoting exist at the empty transition $\varepsilon$.
For example, $(\lambda x:\I.x)\ (1+2)$ is at $\varepsilon$ transition and 
$\TB_\alpha (\lambda x:\I.x)$ is at $\alpha$ transition.
Terefore, transitions appear in typing rules, too.
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A}{\TTB} \andalso
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}}{\TTBL}
\end{center}
\TTB, corresponding to brackets, means 
if $M$ is typed $\tau$ at transition $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
\TTBL\ is converse of \TTB.

@@

CSP, cross-stage persistence, is an important feature of \LTP.
It enables us to embed value at an outer transition into an inner transition.
$\%$ is dedicated to CSP in \LTP.
\AI{What should me mentioned here is that in MetaOCaml CSP is implicit but
in \LTP it's explicit.}
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$
\red{inner / outer は適切か?}
\AI{What is \(a\)?}
\AI{Is this explanation enough to understand the rest of the section?}

@@

Next, we develop \LMD by extending \LTP with \LLF-like dependent types.
From here, we use the word "stage" instead of "transition" 
because we develop a multi-stage calculus, not a multi-transition calculus.
\red{stageのほうが言葉としてふさわしいと言いたい}
There are three technical points in the extension from \LTP to \LMD.

>>>>

\section{Informal Overview of \LMD \label{sec:informal-overview}}

@@

% \red{\LTP\cite{Hanada2014}が頻発するが同一論文の複数回引用に何かルールはあるのか?}
% \AI{ないです．たしかにあんまりつけるとうるさいですね．2.1 の冒頭とかは文ごと不要なのでは？}
\subsection{\LTP}

@@

In \LTP, brackets and escape are written $\TB_\alpha M$ and $\TBL_\alpha M$, respectively.
The type of $\TB_\alpha M$ is $\TW_\alpha \tau$ if $M$ has type $\tau$.
The type of $\TBL_\alpha M$ is $\tau$ if $M$ has $\TW_\alpha \tau$.
% Please notice that if $\TBL_\alpha M$ is well-typed, $M$ has a code type from the typing rule of \LTP.
In addition to normal $\beta$-reduction, there is a reduction rule for brackets and escape.
\begin{align*}
	\TBL_\alpha (\TB_\alpha M) \longrightarrow M 
\end{align*}
It means escape cancels brackets.
% This reduction is called $\longrightarrow_\Lambda$ in \LMD.
% \red{この文はここに来るべきなのか? LTPの説明ではないが}

% stage variable and stage

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage variable}
\footnote{In the original \LTP paper, this is called a \textit{transition variable}. 
However, we call it a \textit{stage variable} in this paper.} and
a sequence of stage variables is called \textit{stage}.
The empty stage is represented by $\varepsilon$.
They are used to show the thickness of brackets. \AI{Perhaps ``thickness'' doesn't make sense.}
For example, $\TB_\alpha (\lambda x:\I.x+10)$ is the fragment of code which becomes $(\lambda x:\I.x+10)$ after it was run once and
$\TB_\alpha \TB_\beta (\lambda x:\I.x+10)$, which is abbreviated as $\TB_{\alpha\beta} (\lambda x:\I.x+10)$,
becomes $(\lambda x:\I.x+10)$ after it was run it twice.
\red{run twiceは表現が微妙}

% stage abstraction and application

There are abstractions for stage variables and applications for stage abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M A$, respectively.
A stage abstraction binds a stage variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
There is a reduction rule for stage application in \LTP.
The rule is the following.
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
For example, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ (\beta\gamma)$ reduces to $\TB_{\beta\gamma} (\lambda x:\I.x)$.

% stage-related symbols disappear when the empty stage is substituted.
% In short, about "run"

Another important rule about a stage variable is 
that symbols with stage variables disappear 
when the empty stage $\varepsilon$ is substituted to the stage variable.
The purpose of this rule is express \verb|run| without any special symbol.
In multi-stage calculus, \verb|run| is a very important operator which changes quoted code to the original code.
For example, $(\TB_\varepsilon (\lambda x:\I.x))$ is equivalent to $(\lambda x:\I.x)$.
Therefore, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ \varepsilon$ becomes $\lambda x:\I.x$
In \LTP, \verb|run| is realized with application to the empty stage $\varepsilon$.

% stage in typing rules

A type judgement of \LTP is of the form $\G \vdash M : \tau @ A$.
A stage in a judgement represents where the judgement is true.
For example, $\G \vdash (\lambda x:\I.x) : \I \to \I @ \alpha$ means 
term $\lambda x:\I.x$ has type $\I \to \I$ at stage $\alpha$.
Especially, terms without quoting exist at the empty stage $\varepsilon$.
For example, $(\lambda x:\I.x)\ (1+2)$ is at $\varepsilon$ stage and 
$\TB_\alpha (\lambda x:\I.x)$ is at $\alpha$ stage.
Terefore, stages appear in typing rules, too.
\begin{center}
	\infrule{\G\vdash M:\tau @{A\alpha}}{\G\vdash \TB_{\alpha}M:\TW_{\alpha}\tau @A}{\TTB} \andalso
	\infrule{\G\vdash M:\TW_{\alpha}\tau @A}{\G\vdash \TBL_{\alpha}M:\tau @{A\alpha}}{\TTBL}
\end{center}
\TTB, corresponding to brackets, means 
if $M$ is typed $\tau$ at stage $A\alpha$ then $\TB_{\alpha}M$, quoted $M$, is typed $\TW_{\alpha}\tau$ at $A$.
\TTBL\ is converse of \TTB.

@@

CSP, cross-stage persistence, is an important feature of \LTP.
It enables us to embed value at an outer stage into an inner stage.
$\%$ is dedicated to CSP in \LTP.
\AI{What should me mentioned here is that in MetaOCaml CSP is implicit but
in \LTP it's explicit.}
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$
\red{inner / outer は適切か?}
\AI{What is \(a\)?}
\AI{Is this explanation enough to understand the rest of the section?}

@@

Next, we develop \LMD by extending \LTP with \LLF-like dependent types.
There are three technical points in the extension from \LTP to \LMD.
